CrypTool Script \\ Mathematics and Cryptography

### 2002-06-26 script/en/chapters/numbertheory.tex
clean up
-- -> -
misc

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\subsection{Examples of modular calculations}

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid’s extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{Appendix_A}{appendix 
A}). Alternatively, we can also use the primitive roots. }.

@@

\subsection{Euler function, Fermat's Little Theorem and Euler--Fermat}

@@

The problems of \index{Factorising} factorising $n=pq$ with $q\neq p$ and finding $J(n)$ have a similar degree of difficulty 
and if we find a solution for one of the two problems, we also have a solution for the
other\footnote{If we know the factors of $n=p*q$ with $p\neq q$, then $J(n)=(p-1)*(q-1) = n-(p+q)+1$. Additionally the factors $p$ and $q$ are
solutions of the quadratic equation
$ x^2 - (p+q) x + pq=0.
$
If only $n$ and $J(n)$ are known, then it is:
$pq=n$ and $p+q= n-J(n)+1.$ So you get $p$ and $q$ by solving the equation
$$ x^2 + (J(n)-n-1)x +n= 0.
$$} (please compare requisition 3 in section 3.10.1).

@@

\subsubsection*{Values of $a^i{\rm ~mod~}46, 1\leq a,i<23$:}
{ \textmd \small
%\begin{tabular}{|p{16 pt}||p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|p{7 pt}|c|}
\begin{center}
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & 1  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean something completely different.} 
procedure in the group $\mathbb{Z}_n^*$.

@@

almost expired.} \footnote{If the two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is not true for all $ m<n$ (although

@@

                be too small.} \footnote{We start by determining either $d$ or $e$ depending on the implementation.}. We can then ``throw away'' $J(n)$.

@@

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice):
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
Such number-theory tasks can also be solved using other tools such as the LiDIA\index{LiDIA} or BC package (see URLs in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.}
\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1; MultiplicativeOrder[11, 347] = 173}
\vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +6 pt
} is extremely difficult to solve.

@@

\bibitem[Pfleger1997]{Pfleger1997} \index{Pfleger 1997}
    Charles P. Pfleger, \\
    Security in Computing, Prentice-Hall, 2nd edition, 1997.

@@

\newpage
\subsection*{URLs / links (selection)}\addcontentsline{toc}{subsection}{URLs}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\pagebreak
\subsection{Examples of modular calculations}

\begin{center}
\fbox{\parbox{15cm}{
    \emph{Seneca\footnotemark:}\\
    The way of theory is long, it is short and effective by examples.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  Lucius Annaeus, philosophical writer and poet, 4 B.~C. $-$ 65 A.~D.
}

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid’s extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{Appendix_A}{appendix 
A}). Alternatively, we can also use the primitive roots. }.

@@

\subsection{Euler function, Fermat's Little Theorem and Euler-Fermat}

@@

The problems of \index{Factorising} factorising\index{Factorisation problem} $n=pq$ with $q\neq p$ and finding $J(n)$ have a similar degree of difficulty 
and if we find a solution for one of the two problems, we also have a solution for the
other\footnote{If we know the factors of $n=p*q$ with $p\neq q$, then $J(n)=(p-1)*(q-1) = n-(p+q)+1$. Additionally the factors $p$ and $q$ are
solutions of the quadratic equation
$ x^2 - (p+q) x + pq=0.
$\\
If only $n$ and $J(n)$ are known, then it is:
$pq=n$ and $p+q= n-J(n)+1.$ So you get $p$ and $q$ by solving the equation
$$ x^2 + (J(n)-n-1)x +n= 0.
$$} (please compare requisition 3 in section 3.10.1).

@@

\subsubsection*{Values of $a^i{\rm ~mod~}46, 1\leq a,i<23$:}
{ \textmd \small
\begin{center}
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & 1  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
Within literature and movies not only cassical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle), but also the RSA scheme. Examples are
\begin{itemize}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner, 1999
  (included in the CrypTool package as PDF file), where the sisters
  privately communicate using a variant of RSA.
\end{itemize}
} 
procedure in the group $\mathbb{Z}_n^*$.

@@

almost expired.}$^,$\footnote{If the two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is not true for all $ m<n$ (although

@@

                be too small.}$^,$\footnote{We start by determining either $d$ or $e$ depending on the implementation.}. We can then ``throw away'' $J(n)$.

@@

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice):
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.}
\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1; MultiplicativeOrder[11, 347] = 173}
\vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +6 pt
} is extremely difficult to solve.

@@

\bibitem[Pfleeger1997]{Pfleeger1997} \index{Pfleeger 1997}
    Charles P. Pfleeger, \\
    Security in Computing, Prentice-Hall, 2nd edition, 1997.

@@

\newpage
\subsection*{Web links}\addcontentsline{toc}{subsection}{Web links}
